# R is Friend

::: callout
## Learning Objectives

+ Identify three of the foundational data types in R
+ Explain similarities and differences between different object types
+ Use indexing to extract elements from different object types
+ Use functions to calculate summary statistics
+ Use logical values and indexing to calculate summary statistics
:::

## Data Types

In the last class, you learned how to create simple objects (e.g., `x <- 1221`, 
`name <- "jane"`), but we never talked about the difference between a value 
like `1221` and `"jane"`. These two values have different **data types**---one
is a **number** and one is a **character** (really, a set of characters). These 
are two of what are called the ["atomic" data types in R](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Vectors). Let's 
dig into these data types more!

### Numeric Values

You've seen a lot of numbers in your day, so it probably wouldn't surprise you
to know that there are different types of numbers. There are numbers **with**
decimals and there are numbers **without** decimals. Technically, R separates
these two types of numbers into their own data types---**doubles** and
**integers**.

#### Double (Real) Values

If a number has a decimal value (e.g., `26.2`), then R will store the number 
as a double (real) numeric. 

```{r}
#| echo: true
#| eval: true
#| label: example-make-double

room <- 26.2
typeof(room)
```

::: {.callout-tip}
# `typeof()`

The `typeof()` function tells you the storage type that R is using internally to
hold the data. You can think of this function asking the question, 
"How is this data physically stored in memory?". 
:::

#### Integer Values

Numbers that **do not** have decimals are called `integer`s. If you want a number
to be stored as an `integer`, you need to tell R to do so. Otherwise, your
number will still be stored as a `double`.

::: columns
::: {.column width="48%"}

Defaults to `double`:

```{r}
#| echo: true
#| eval: true
#| label: example-make-integer-double

room <- 1221
typeof(room)
```

:::

::: {.column width="4%"}
:::

::: {.column width="48%"}
Adding an `L` to the end creates an `integer`:

```{r}
#| echo: true
#| eval: true
#| label: example-make-integer-1

room <- 1221L
typeof(room)
```
:::
:::

::: {.callout-opinion}

You might wonder why `L` is used to create an `integer`. Wouldn't an `I` make 
more sense?

Because R is written mostly in C (another programming language), it likely
inherits some naming conventions from C. In C, `L` is used to denote a (long)
integer. Moreover, it would be confusing to use an `I` to indicate an `integer`
since you use a `"i"` suffix to create a complex number. 

:::

::: {.callout-tip}
# Other ways to create an `integer`

Adding an `L` at the end of a number is one way to make an integer, but this
really only works when we are creating our objects by hand. More frequently, 
we will be given a vector of numbers which we may want to transform into 
integer values. The `as.integer()` function can help us do just that!

```{r}
#| echo: true
#| eval: true
#| label: example-make-integer-2

room <- as.integer(1221)
typeof(room)
```
:::

<!-- We could choose to highlight how `class()` returns the same value (numeric) for -->
<!-- both doubles and integers.  -->

<!-- class() tells you the high-level type that R uses to decide how to behave with -->
<!-- the object. Think of it as asking: "What role does this object play?"  -->

<!-- Imagine a chef's knife: -->
<!-- typeof() → "It's made of stainless steel" (the raw material) -->
<!-- class() → "It's a kitchen tool" (the role it plays) -->

### Character Values

### Logical Values

```{r}
passed <- TRUE
typeof(passed)
```

### Other Data Types

So far, we've covered real (double), integer, character, and logical data types.
These form all but two of the atomic data types. The only two that remain are
**complex** and **raw**. As you might expect from its name, the complex data
type is used to store complex numbers (e.g., `4 + 3i`). The raw data type is 
possibly the least used data type, whose purpose is to store data at the byte
level.

In STAT 1810 and 3820 we will **only** work with real, integer, character, and
logical data types!

## Object Types
<!-- Intro to vectors, lists, matrices, dataframes, tibbles.  -->
<!-- Building up objects -->
<!-- Small object (from last week) is a scalar value -->
<!-- Bigger object is a vector (multiple values) -->
<!-- What if we have multiple vectors? -->
<!-- To store them in a matrix, they all need to be the same data type. -->
<!-- If they are different data types, then we need a list.  -->
<!-- What if the vectors are all the same length? -->
<!-- Then we can use a data frame (or a tibble)! -->

<!-- In the lab they learned about using the c() function to make a vector. -->
<!-- The last problem explored datatype coercion when a string is added to a vector of numbers -->

### Vectors
<!-- Give basic info / examples. -->
<!-- How to create each type of structure. -->


### Multiple Vectors
<!-- Relate structures to each other – e.g., a column of a matrix / dataframe / tibble is a vector. -->

#### Do all of the vectors have the same data type?
<!-- matrices! -->


#### Do the vectors have different data types?
<!-- lists! -->

#### Are the vectors all the same length?
<!-- dataframes or tibbles! -->


::: {.callout-check-in}
<!-- How to recognize what each type looks like printed out. -->

:::

## Indexing Objects
<!-- Indexing and subsetting -->
<!-- Accessing columns in a dataframe / tibble via $. -->
<!-- Accessing elements via []. -->
<!-- Accessing elements via T / F. -->

### Vectors

### Matrices

### Lists

### Dataframes


::: {.callout-check-in}
<!-- Give them data / coding scenarios and they pick which structure would be
most appropriate and then code a small example. -->

:::


## Mathematical Operations
<!-- Can do math via indexing – add 1 to the first element, then the second element, etc. -->
<!-- Vectorized math. -->
<!-- But many functions are vectorized -- +, -, x. /, exp, log, ==, etc. -->

### Operations on Columns

### Operations on Rows

### Operations with Logical Values
<!-- Finding sums / proportions via T / F. -->
