# Functions

::: callout
## Learning Objectives
+ 
+ 
+
:::

<!-- Functions they've seen so far: -->
<!-- typeof() -->
<!-- as.character() -->
<!-- as.numeric -->
<!-- c() -->
<!-- matrix -->
<!-- list  -->
<!-- data.frame -->
<!-- seq() -->
<!-- sqrt() -->
<!-- length() -->
<!-- sum() -->
<!-- case_when() -->
<!-- print() -->

At this point, you've encountered almost 20 functions either built-in to R 
(e.g., `sqrt()`) or that live in specific packages (e.g., `case_when()`). For
each function, we've learned how we can use it to get the results we were after, 
but we haven't dug into what makes a function a function. That's what this 
reading is about!

## Motivation

It is common in statistics and data science to need to rescale a variable before
including it in a model. Rescaling is quite helpful when variables are on 
very different scales. In fact, for some machine learning methods (e.g., 
K-nearest neighbors) it is critical that **every** variable be on the same 
scale for an analysis. 

Take for example, some common measures of the economy. The consumer price index
(CPI) measures the yearly change in price for a representative basket of goods
and services[^1] and is recorded as a weighted average of prices (e.g., $250). 
Alternatively, the PPI (producer price index) measures the yearly change in
prices paid to domestic producers of goods and services (e.g., $120). Finally, 
the PCE (Personal Consumption Expenditures) is a slightly broader measure than
CPI in that it attempts to measure both how much households are spending 
(e.g., $14,000) and what they're spending their money on. 

[^1]: [The Bureau of Labor Statistics (BLS)](https://www.bls.gov/cpi/questions-and-answers.htm)
has a database of over 80,000 prices which they sample monthly for more than 200
categories of products and services to determine the CPI and inflation rate. The
BLS uses a huge basket because its goal is to get an accurate measure of price
changes for consumer goods and services across the U.S. economy. 

<!-- Data obtained from https://fred.stlouisfed.org/ -->

::: columns
::: {.column width="23%"}
```{r}
#| label: economy-year

year <- c(2020, 
          2021, 
          2022, 
          2023, 
          2024, 
          2025)
```
:::

::: {.column width="2%"}
:::

::: {.column width="23%"}
```{r}
#| label: economy-cpi

cpi <- c(258.811, 
         270.970, 
         292.655, 
         304.702, 
         313.689, 
         321.943)
```
:::

::: {.column width="2%"}
:::

::: {.column width="23%"}
```{r}
#| label: economy-ppi

ppi <- c(119.24, 
         119.46, 
         127.82, 
         139.98, 
         142.80, 
         146.16)
```
:::
::: {.column width="2%"}
:::

::: {.column width="25%"}
```{r}
#| label: economy-pce

pce <- c(14231.33,
         16119.67,
         17690.02,
         18833.15,
         19896.01,
         20954.86)
```
:::
:::

Since we have economic measures for each year[^2], let's assemble them into a
dataframe:

[^2]: If we have vectors that are the **same** length, then each vector can be
a column of a dataframe. If they were *different* lengths we would need to use
a list to store them all together!

```{r}
economy <- data.frame(year = year, 
                      cpi = cpi, 
                      ppi = ppi, 
                      pce = pce)

economy
```

Notice how each of these variables are on different scales? Well, that's exactly
why we might want (need) to rescale them! 

### Min-Max Scaling 

One common method for rescaling is called "min-max scaling." For any numeric
vector, this method finds the maximum value and the minimum value and then uses
these values to rescale each value of the vector. Mathematically, each $i^{th}$ 
value of a vector $x$ is scaled based on this calculation:

$$\frac{x_i - min(x)}{max(x) - mix(x)}$$
For example, if we have a vector `x` with values `c(1, 2, 5, 7)`. Then the
minimum of `x` is `1` and the maximum of `x` is `7`. We then use these values
to rescale every value of `x`. This would look like the following code:

```{r}
#| label: example-min-max

x <- c(1, 2, 5, 7)

(x - min(x)) / (max(x) - min(x))
```

Notice that the minimum value is now 0 because `1 - 1` (the minimum is 1) is 0
and the maximum value is now 1 because `(7 - 1) / (7 - 1)` (the maximum is 7) is
1. 

We could rescale every column of `economy` using the same code as above:

```{r}
#| label: repeated-rescaling
#| eval: false

economy$cpi <- (economy$cpi - min(economy$cpi)) / (max(economy$cpi) - min(economy$cpi))
economy$ppi <- (economy$ppi - min(economy$ppi)) / (max(economy$ppi) - min(economy$pci))
economy$pce <- (economy$pce - min(economy$pci)) / (max(economy$pce) - min(economy$pce))

```

There are a couple of issues with this process. By repeating this same process 
multiple times, it is harder to understand what the code is doing **and** we've
made our code more error prone. Did you notice the error we made?

A better method would be to write a function that allows us to implement this 
process anytime we want to rescale a numeric vector!

### Don't Repeat Yourself

The critical motivation behind functions is the "don't repeat yourself" (DRY)
principle. In general, "you should consider writing a function whenever copied
and pasted your code **more than twice** (i.e. you now have three copies of the
same code)" (Wickham & Grolemund, 2020).

One of our favorite papers, *Best Practices for Scientific Computing*, 
summarizes this idea in a slightly different way:

> Anything that is repeated in two or more places is more difficult to maintain.
> Every time a change or correction is made, multiple locations must be updated,
> which increases the chance of errors and inconsistencies. To avoid this,
> programmers follow the DRY (Don't Repeat Yourself) Principle, which applies to
> both data and code.

The DRY Principle advocates that you *modularize* your code---break it into 
smaller, independent, and reusable units. Modularizing your code helps you 
understand what your code is doing and makes your code more easily re-purposed
for other projects. 

## Functions


### Syntax


### Arguments + Parameters

<!-- add an argument that allows the user to choose between min/max and mean/sd scaling -->


## Debugging

<!-- print(), run line by line, etc. -->


## Input validation


## Environment, Scope, Name Masking & Dynamic Lookup

<!-- give examples and ask them to identify where the values are pulled from? -->


