# More Wrangling {.unnumbered}


```{r}
#| label: setup
#| echo: false

library(tidyverse)
```


::: callout
## Learning Objectives
+ Determine what data format is necessary to generate a desired plot or statistical model.
+ Use helper functions like `across()` and `case_when()` with dplyr verbs.
+ Create small vector functions for use in wrangling.
+ Describe the differences between “wide” and “long” format data and how to transition between the two structures.
+ Design and implement more advanced plots
:::


Last week, we learned many key tools and skills to wrangle data in R. This week, we will fill out those skills with a couple of additional functions that work assist wtih the primary `dplyr` verbs. We will also consider how to tidy data so that we have the best rows and columns avaiable to complete a desired analysis.


## Extending `dplyr` verbs

```{r}
#| echo: false

# load liver package
library(liver)

# load cereal data which is included in the package
data(cereal)
```

### `if_else()` and `case_when()`

<!--Need to edit, given material in 3a-control flow.-->


You learned about vectorized conditional statements in @sec-3a-case-when. These functions, `if_else()` and `case_when()` are useful helper functions with `mutate()` to create new categorical (character) variables from an existing numeric or character variable.

`if_else()` implements one `if-else` statement with the format:

 `if_else(<CONDITION>, <TRUE OUTPUT>, <FALSE OUTPUT>)`
 
 If the logical condition provided is met, then the `<TRUE OUTPUT>` is returned, otherwise, the `<FALSE OUTPUT>` is returned. Note that `if_else()` does not have a `.data = ` argument -- it must be used within `mutate()`!
 
For example, let's say I want to label each cereal as either low or high potassium:

```{r}
#| echo: true
#| eval: false
cereal |> 
  mutate(po_category = if_else(potass <= 100, 
                               "low", 
                               "high"),
         .after = potass)
```

```{r}
#| eval: true
#| echo: false
cereal |> 
  mutate(po_category = if_else(potass <= 100, "low", "high"),
         .after = potass) |> 
  knitr::kable() |> 
  kableExtra::scroll_box(height = "200px") |> 
  kableExtra::kable_styling(font_size = 30)
```
 
But what if I want more than two categories? Say I want to label each cereal as "low", "medium", "high", or "very high" potassium.


As you learned in @sec-3a-case-when, `case_when()` implements a **series** `if-else` statements one after another, so we can create these categories:

```{r}
#| echo: true
#| eval: false
cereal |> 
  mutate(po_category = case_when(potass == -1 ~ NA_character_,
                                 potass <= 50 ~ "low", 
                                 potass <= 75 ~ "medium",
                                 potass <= 100 ~ "high",
                                 potass > 100 ~ "very high"),
         .after = potass)
```

```{r}
#| eval: true
#| echo: false
cereal |> 
  mutate(po_category = case_when(potass == -1 ~ NA_character_,
                                 potass <= 50 ~ "low", 
                                 potass <= 75 ~ "medium",
                                 potass <= 100 ~ "high",
                                 potass > 100 ~ "very high"),
         .after = potass) |> 
  knitr::kable() |> 
  kableExtra::scroll_box(height = "200px") |> 
  kableExtra::kable_styling(font_size = 30)
```

:::{.callout-tip collapse = "true"}
### Coding missing values

From some data exploration, we noticed that there were sometimes negative values for potassium in the data, which seems weird!

Looking at the data dictionary, it turns out missing values were coded as the number `-1`. This is actually very common in publicly available data.

We wanted to make sure to code `-` as `NA` rather than `"low"` in our new variable. 

A trick in R is that there are technically different "kinds" of `NA`s based on the variable type. Since we were creating a character variable, we needed to use `NA_character_`. You can read more about it in the [documentation](https://rdrr.io/r/base/NA.html).
:::

See @sec-3a-case-when for additional review on the arguments to `case_when()`. As a reminder, `case_when()` will return a missing value for rows that weren't captured by the logical statements provided. The `.default` and `.unmatched` arguments are helpful in this case!


### `across()` 

<!-- maybe a good moment here to use a custom vector function -->



### `pull()`

At this point you should hopefully be noticing that the output of all of the dplyr functions discussed so far are a dataframe or tibble. This is great, and by design, but for some tasks, this might not be ideal.

For example, say I want to calculate the average potassium level for cereals with more than 3 grams of protein. Sounds like I want to first *filter* the observations and then *summarize* the data.

```{r}
cereal |> 
  filter(protein > 3) |> 
  summarize(mean_potass = mean(potass, na.rm = T))
```
This gets me the answer, but the output is a dataframe. What if I just wanted to get a numeric output? 

```{r}
cereal |> 
  filter(protein > 3) |> 
  select(potass) |> 
  mean(na.rm = T)
```

Darn, this doesn't work because the output of `select(potass)` is a **dataframe** with one column. The `mean()` function is expecting a **vector** as an input. What I really want to do is *pull* out the `potass` data variable as a vector! If I wasn't in the middle of a pipeline, I could use `cereal$potass` to index the `potass` column. `pull()` extracts a column from a data frame as a vector, which is very useful within pipelines. 

Now we can fix our faulty code from before:

```{r}
cereal |> 
  filter(protein > 3) |> 
  pull(potass) |> 
  mean(na.rm = T)
```






